Modal operators for a constructive account of well quasi-orders

### 2025-02-25 slides-paris2025.tex
Slides for Logique à Paris 2025, almost finished

\setbeameroption{show notes on second screen=bottom}
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[width=\paperwidth]{tower}\end{minipage}}
\begin{frame}[c]
  \centering
  \color{white}

@@

  \setbeamercolor{block body}{bg=black!100}
  \begin{minipage}{0.62\textwidth}
    \begin{block}{\centering Arbeitstagung Bern--München--Verona}
      \centering\normalsize\color{white}
      \hil{Modal operators for \\ a constructive account of
      well quasi-orders} \\[-0.9em]
      \
    \end{block}
  \end{minipage}

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bf
  \colorbox{black}{November 30th, 2024}%
  \bigskip

@@

  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff~$\mathsf{Good} \mid
  [\,]$, where~$\mathsf{Good}\,x_0 \ldots x_{n-1}$ iff~$\exists i < j\_ x_i \leq x_j$.
  \pause
  \pause

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{staircase-more-faded}\end{minipage}}
\begin{frame}{The modal multiverse of constructive forcing}
  \jnote{1}{
    By \emph{topos}, we mean \emph{Grothendieck topos}. In constructive
    forcing, a ``forcing extension of the base universe'' is exactly the same
    thing as a Grothendieck topos.

@@

  \textbf{Def.} A statement~$\varphi$ holds \ldots
  \begin{itemize}
    \small
    \item \hil{everywhere} \tabto{2.15cm}($\necessary\varphi$)\tabto{3.08cm} iff it holds in every topos
    (over the current base).
    \item \hil{somewhere} \tabto{2.15cm}($\possible\varphi$)\tabto{3.08cm} iff it holds in some positive topos.
    \item \hil{proximally} \tabto{2.15cm}($\xpossible\varphi$)\tabto{3.08cm} iff it holds in some positive overt topos.
  \end{itemize}
  \bigskip

  \begin{tikzpicture}[overlay]
    \node[anchor=south east,inner sep=0] (image) at (14.8,0.6) {
      \includegraphics[width=0.2\textwidth]{branching}
    };
  \end{tikzpicture}\vspace*{-2.3em}
  \pause

  \colorbox{white}{\bignumber{1} A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.}

  \mbox{\qquad
  \colorbox{white}{\bignumber{2} A relation is well-founded iff \emph{everywhere}, there is
  no infinite descending chain.}}

  \qquad
  \qquad
  \colorbox{white}{\bignumber{3} A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.}

  \mbox{\qquad
  \qquad
  \qquad
  \colorbox{white}{\bignumber{4} For every inhabited set~$X$, \emph{proximally} there is an
  enumeration~$\NN \twoheadrightarrow X$.}}

  \qquad
  \qquad
  \qquad
  \qquad
  \colorbox{white}{\bignumber{5} For every ring, \emph{proximally} there is a maximal ideal.}

  \qquad
  \qquad
  \qquad
  \qquad
  \qquad
  \colorbox{white}{\bignumber{6} \emph{Somewhere}, the law of excluded middle holds.}

  \pause
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-4.68em}\includegraphics[width=1.2\paperwidth]{wqo-faded}\end{minipage}}
\begin{frame}{Answering a question by Berardi--Buriola--Schuster}
  \jnote{1-3}{
    Stefano Berardi, Gabriele Buriola and Peter Schuster recognized in their
    recent article
    \href{https://drops.dagstuhl.de/entities/document/10.4230/LIPIcs.CSL.2024.16}{A
    general constructive form of Higman's lemma} the displayed implication as an
    efficient organizing principle for structuring a constructive account of
    well quasi-orders, which they elevated to the status of a definition.

    They recognized that their definition implies well\ind (and hence
    well\sinf), and posed the converse as an open question.

    The generating clauses for~$B \mid_\text{incr} \sigma$, for increasing
    finite lists~$\sigma$, are the following:

    \vspace*{-0.5em}
    \begin{enumerate}
      \item If~$B\sigma$, then~$B \mid_\text{incr} \sigma$.
      \\[-1.3em]
      \item If~$B \mid_\text{incr} \sigma x$ for all~$x \in X$ such that~$\sigma x$ is
      increasing, then~$B \mid_\text{incr} \sigma$.
    \end{enumerate}
  }

  \jnote{4}{
    The modal language unlocked the displayed positive answer to their
    question. As the modal framework is fully constructive, the proof can in
    principle be unrolled to yield a direct (but likely longer) argument not
    involving constructive forcing, giving an explicit algorithm of type
    \[ \mathsf{Good} \mid [\,] \ \times\ B \mid_\text{incr} [\,]
    \ \longrightarrow\ B \mid [\,] \]
    for transforming given witnessing trees. This unwinding is work in
    progress.
  }

  \textbf{Def.} A quasi-order~$X$ is \hil{well\impl} iff (approximately) for
  every monotone predicate~$B$,
  \[ \text{if $B \mid_{\text{incr}} [\,]$, then $B \mid [\,]$.} \]
  \justifying
  {\small``Assume that no matter how the empty list evolves to an \emph{increasing}
  list~$\sigma$, eventually~$B\,\sigma$. Then no matter how the empty
  list evolves to an \emph{arbitrary} list~$\tau$, eventually~$B\,\tau$.''\par}
  \pause

  \hil{Equivalently:} If everywhere every \emph{increasing} infinite sequence~$\alpha : \NN
  \to X$ has a finite prefix validating~$B$, then so does every
  \emph{arbitrary} infinite sequence everywhere.
  \pause

  \textbf{Prop.} $\text{well\impl} \Rightarrow \text{well\ind}$.

  \emph{Proof.} Trivially~$\mathsf{Good} \mid_{\text{incr}} [\,]$,
  hence~$\mathsf{Good} \mid [\,]$ by assumption. \qed
  \pause

  \textbf{Thm.} $\text{well\impl} \Leftarrow \text{well\ind}$.

  \small
  \emph{Proof.} Let~$x_0,x_1,\ldots$ be an infinite sequence (in an arbitrary topos).
  \emph{Somewhere,} \textsc{lem} holds and implies that there is an increasing infinite subsequence~$x_{i_0},
  x_{i_1}, \ldots$. \emph{There} we have, by assumption, a finite prefix
  with~$B\,x_{i_0}\ldots x_{i_n}$. As~$B$ is monotone, \emph{there}
  we also have~$B\,x_0x_1\ldots x_{i_n}$. So \emph{somewhere} there is a finite
  prefix validating~$B$. Hence there actually is a finite prefix
  validating~$B$. \qed
\end{frame}}

\backupstart
{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-2.2em}\includegraphics[width=\paperwidth]{agda-wqo}\end{minipage}}
\begin{frame}
  \vspace*{\paperheight}\vspace*{-2.5em}\centering
  \emph{Agda formalization in progress.}
\end{frame}}
\backupend

@@

>>>>

%\setbeameroption{show notes on second screen=bottom}
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}

@@

\definecolor{mypurpleblack}{RGB}{30,0,50}
\setbeamercolor{structure}{fg=black}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-7em}\includegraphics[width=\paperwidth]{forest2}\end{minipage}}
\begin{frame}[c]
  \centering
  \color{white}

@@

  \setbeamercolor{block body}{bg=black!100}
  \begin{minipage}{0.52\textwidth}
    \begin{block}{}
      \centering\normalsize\color{white}
      \hil{\color{white}Computing an integer using the \\ modal toposophic multiverse} \\[-0.4em]
      \
    \end{block}
  \end{minipage}

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bf
  \colorbox{black}{\begin{minipage}{0.2\textwidth}
    \centering
    Logique à Paris \\
    February 26th, 2025
  \end{minipage}}
  \bigskip

@@

  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff there exists a
  \hil{modulus of wellness} for~$X$.
  \pause
  \pause

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{0.29cm}\hspace*{-1cm}\includegraphics[height=\paperheight]{tower-cropped}\end{minipage}}
\begin{frame}{The modal multiverse of constructive forcing}
  \jnote{1}{
    By \emph{topos}, we mean \emph{Grothendieck topos}. In constructive
    forcing, a ``forcing extension of the base universe'' is exactly the same
    thing as a Grothendieck topos.

@@

  \begin{changemargin}{7.5em}{-0.5em}
    \textbf{Def.} A statement~$\varphi$ holds \ldots
    \vspace*{-0.3em}
    \begin{itemize}
      \small
      \item \hil{everywhere} \tabto{2.15cm}($\necessary\varphi$)\tabto{3.08cm}
      iff it holds \hil{in every topos}
      (over the current base).
      \item \hil{somewhere} \tabto{2.15cm}($\possible\varphi$)\tabto{3.08cm} iff
      it holds \hil{in some positive topos}.
      \item \hil{proximally} \tabto{2.15cm}($\xpossible\varphi$)\tabto{3.08cm}
      iff it holds \hil{in some positive overt topos}.
    \end{itemize}

    \textbf{Def.} A (Grothendieck) \hil{topos} is a category equivalent to the
    category of sheaves over a small site.

    \textbf{Examples for toposes.}
    \begin{itemize}
      \item $\mathrm{Set}$, the category of sets and maps.
      \item The category of sets and maps which are \hil{defined up to~$\boldsymbol\neg\neg$}.
      \item $\mathrm{Set}[G]$, the extension obtained by adding a \hil{generic
      filter} of a \hil{forcing notion} (a quasi-order equipped with a
      coverage).
    \end{itemize}
    The following are \bad{not} toposes:
    \begin{itemize}
      \item The category of sets and partially defined maps.
      \item The category of abelian groups.
    \end{itemize}

    \begin{enumerate}
      \item A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.
      \medskip
      \item A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.
      \medskip
      \item For every inhabited set~$X$, \emph{proximally} there exists an
      enumeration~$\NN \twoheadrightarrow X$.
      \medskip

      \item For every ring, \emph{proximally} there exists a maximal ideal.
      \medskip

      \item \emph{Somewhere}, the law of excluded middle holds.
    \end{enumerate}
  \end{changemargin}
\end{frame}}

@@

- Well quasi-orders
  *abstract proofs as blueprints for concrete computations?*
  - def. good, well
  - ex. ℕ
  - classical stability results

- A better notion
  - partially defined sequences
    (not really a morphism in Set, but in Sh_¬¬(Set))
  - multivalued sequences
    (approximately morphisms in some category)
  - modulus of wellness

- The modal toposophic multiverse
  - examples for toposes:
    - Set
    - Sh_¬¬(Set)
    - Set[G], obtained by adding a generic filter of a forcing notion
    - not quite toposes: arbitrary partially defined maps, abelian groups
  - modal operators
  - Diamond φ ⇒ φ etc.
  - X has a modulus of wellness iff everywhere, every sequence is good
  - Diamond (¬LEM)
  - etc.

- Applications
  - Dickson, constructively
  - surjective matrices (with Agda)

Somewhere: eigenvalues etc.
